Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-07-24 paper/RefereeResponses1.tex

\noindent C30) L316: So why do you show values of $\sigma >6$? And also maybe highlight this in the figure Figure 1d and e: I’m semi convinced by these distributions showing ATA. In 1d it simply appears that the variance scales with the density. 1e is quite unconvincing to me. That’s generally fine, but figure 1 seems to be a conceptual figure, so I wonder whether you don’t have a better example? \\

\noindent ***\emph{Regarding the first referee question listed above, well, what we are 
trying to do here is accurately portray our results about
when ATAs were important and when they were not. To us, the most forthright way of doing that is to
show \textbf{all} the results, including the values of $\sigma$ which are probably unrealistic, and mention
(as we do) that some values are unrealistic. The point of this paragraph is to outline the circumstances
we have explored, using our lottery-model approaches, for which ATAs were important or not, and including all
our results seems important when such is our goal. Furthermore, there's no real cost to showing
all the $\sigma$'s we explored, it does not take up more manuscript space to do so.
Some readers may also take the lesson, for future work, from these explorations that one of the things that probably 
matters for continuing to explore when ATAs are important and when not is probably the strength
of the noise impacting the system. Finally, whether a value of $\sigma$ is ``unrealistic'' is 
a bit of judgement call rather than a formal cutoff, to we hesitate to censor results 
based on such a thing.}
%DONE

\emph{Regarding the second part of the referee comment, pertaining to Figure 1d and e, it is perplexing to us that 
this is mentioned here,
as the paragraph at L316 is not directly related to Fig. 1d-e. That figure
shows motivational empirical examples, distinct from the lottery model work discussed around L316. Nevertheless, we took
seriously the referee's comment that they were apparently not convinced there are ATAs in Fig. 1d-e.
In fact, one can significance test these results and they are solidly significant - we added
those results, along with the methods used to get them, to a new SI section which is now cited from the
figure caption. We also added some text to the figure caption which we hope will focus the 
readers attention on the aspects of these panels which clearly show asymmetries of tail association. } \\

### 2023-07-24 paper/Paper.Rnw
made changes to various files to address the referee desire to know that Fig 1d,e are significantly asymmetric

#place R code here for loading in necessary variables from Results
res_loc <- "../results_numeric/"
plankStats <- readRDS(file=paste0(res_loc, "plankStats.RDS"))
#fig2 <- readRDS(file=paste0(res_loc,"fig2data.RDS"))
M <- readRDS(file=paste0(res_loc,"M.RDS"))
se3 <- readRDS(file=paste0(res_loc,"fig3maxse.RDS"))
se4 <- readRDS(file=paste0(res_loc,"fig4maxse.RDS"))
se5 <- readRDS(file=paste0(res_loc, "fig5maxse.RDS"))
#se6 <- readRDS(file=paste0(res_loc, "fig6maxse.RDS"))
@

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig1_a-h.pdf}
\caption{Examples motivating our study of the effects of asymmetric tail associations (ATAs) on competition
and coexistence. (a-c) Bivariate random variables with standard normal marginals showing
alternative patterns of association in the left and right parts of the distributions. 
The two variables in (a) are more strongly related to each
other in the left parts of the distributions, termed ``left-tail association;'' those in (b) 
are symmetrically associated; and those in (c) 
are more strongly related to each other in the right parts, termed ``right-tail association''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these 
very different patterns of association. For (a, c), we used asymmetrically tail associated cases which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also possible to generate asymmetrically 
tail associated noise with imperfect correlations in both halves of the distributions [\citealp{Nelsen2006,Ghosh_2020_AER};
see also (d,e)]. 
The variables $(B_1,B_2)$ described in the Introduction and Methods were obtained by generating noise as 
pictured and then transforming the marginals as specified in Methods.
(d,e) Two examples of ATAs found in nature. Axes are population densities of two plankton species, 
\emph{Ceratium fusus} and \emph{Ceratium furca}, sampled in successive years in two of the 15 locations 
in the North Sea considered by \citep{Ghosh_2020_ecolevol}. ATAs differed by location: 
(d) has a partial Spearman correlation of  \Sexpr{round(plankStats$d,3)}, indicating left-tail association
 and (e) has a partial Spearman correlation of  \Sexpr{round(plankStats$e,3)}, indicating right-tail association \cite{Ghosh_2020_AER}.
(f-h) Lottery model population simulations using $(B_1,B_2)$ obtained by transforming the 
respectively corresponding variables (a-c; see Introduction).
\label{fig:pedagog1}}
\end{figure}

>>>>

#place R code here for loading in necessary variables from Results
res_loc <- "../results_numeric/"

#fig2 <- readRDS(file=paste0(res_loc,"fig2data.RDS"))
M <- readRDS(file=paste0(res_loc,"M.RDS"))
se3 <- readRDS(file=paste0(res_loc,"fig3maxse.RDS"))
se4 <- readRDS(file=paste0(res_loc,"fig4maxse.RDS"))
se5 <- readRDS(file=paste0(res_loc, "fig5maxse.RDS"))
#se6 <- readRDS(file=paste0(res_loc, "fig6maxse.RDS"))
@

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig1_a-h.pdf}
\caption{Examples motivating our study of the effects of asymmetric tail associations (ATAs) on competition
and coexistence. (a-c) Bivariate random variables with standard normal marginals showing
alternative patterns of association in the left and right parts of the distributions. 
The two variables in (a) are more strongly related to each
other in the left parts of the distributions, termed ``left-tail association;'' those in (b) 
are symmetrically associated; and those in (c) 
are more strongly related to each other in the right parts, termed ``right-tail association''. All three cases have 
the same Pearson correlation, $\rho$, up to sampling variation, so Pearson correlation does not identify these 
very different patterns of association. For (a, c), we used asymmetrically tail associated cases which are perfectly 
correlated in the lower or upper halves of the distributions, though it is also possible to generate asymmetrically 
tail associated noise with imperfect correlations in both halves of the distributions [\citealp{Nelsen2006,Ghosh_2020_AER};
see also (d,e)]. 
The variables $(B_1,B_2)$ described in the Introduction and Methods were obtained by generating noise as 
pictured and then transforming the marginals as specified in Methods.
(d,e) Two examples of ATAs found in nature. Axes are population densities of two plankton species, 
\emph{Ceratium fusus} and \emph{Ceratium furca}, sampled in successive years in two of the 15 locations 
in the North Sea considered by \citep{Ghosh_2020_ecolevol}. ATAs were statistically
significant and differed by location, with associations being primarily in the left tails in 
panel d, and primarily in the right tails in panel e (SI section \ref{SI-SIsect:planktonstats}). 
One can see these asymmetries by noting that the open points in panel d, which are in the left halves of the
distributions, are more tightly clustered around the line $y=x$ than are the closed points,
which are in the right halves of the distributions; and \emph{vice versa} for panel e.
(f-h) Lottery model population simulations using $(B_1,B_2)$ obtained by transforming the 
respectively corresponding variables (a-c; see Introduction).
\label{fig:pedagog1}}
\end{figure}

### 2023-07-24 paper/SupMat.Rnw
made changes to various files to address the referee desire to know that Fig 1d,e are significantly asymmetric

\newcommand{\olr}{\overline{r}}
\newcommand{\olrs}{\overline{r}^{\shp}}
\newcommand{\olrn}{\overline{r}^{\nat}}
\newcommand{\olep}{\overline{\epsilon}}
\newcommand{\olE}{\overline{E}}
\newcommand{\olC}{\overline{C}}

@@

#place R code here for loading in necessary variables from Results
#ordinarily this would be a readRDS command, but I'm just defining a variable here for now
#allregres<-matrix(rnorm(10),2,5)
M <- readRDS(file="../results_numeric/M.RDS")
rho <- readRDS(file="../results_numeric/rho.RDS")
se3qij <- readRDS(file="../results_numeric/fig3qijmaxse.RDS")
se4qij <- readRDS(file="../results_numeric/fig4qijmaxse.RDS")
@

@@

>>>>

\newcommand{\olr}{\overline{r}}
\newcommand{\olrs}{\overline{r}^{\shp}}
\newcommand{\olrn}{\overline{r}^{\nat}}
\newcommand{\olep}{\overline{\epsilon}}
\newcommand{\olE}{\overline{E}}
\newcommand{\olC}{\overline{C}}
\newcommand{\cor}{\text{cor}}

@@

#place R code here for loading in necessary variables from Results
#allregres<-matrix(rnorm(10),2,5)

M <- readRDS(file="../results_numeric/M.RDS")
rho <- readRDS(file="../results_numeric/rho.RDS")
se3qij <- readRDS(file="../results_numeric/fig3qijmaxse.RDS")
se4qij <- readRDS(file="../results_numeric/fig4qijmaxse.RDS")

res_loc<-"../results_numeric/"
panD_stat <- readRDS(file=paste0(res_loc, "DiffPartialSpearmanPanelD.Rds"))
panE_stat <- readRDS(file=paste0(res_loc, "DiffPartialSpearmanPanelE.Rds"))
panD_pval <- readRDS(file=paste0(res_loc, "DiffPartialSpearman_pval_PanelD.Rds"))
panE_pval <- readRDS(file=paste0(res_loc, "DiffPartialSpearman_pval_PanelE.Rds"))
panD_parm <- readRDS(file=paste0(res_loc, "DiffPartialSpearman_parm_PanelD.Rds"))
panE_parm <- readRDS(file=paste0(res_loc, "DiffPartialSpearman_parm_PanelE.Rds"))
nsims<-readRDS(file=paste0(res_loc,"DiffPartialSpearmannsims.Rds"))
@

@@

\section{Statistical significance of the plankton example of ATAs of main text Fig. \ref{MT-fig:pedagog1}}\label{SIsect:planktonstats}

For scatterplot data showing positive associations, such as those pictured in Fig. \ref{MT-fig:pedagog1}d-e, 
\cite{Ghosh_2020_AER} defined an index which takes positive values when association
is primarily in the left halves of the distributions, and negative values when association
is primarily in the right halves of the distributions. See below for a restatement of the definition
of the index of \cite{Ghosh_2020_AER}. Values of the index were \Sexpr{round(panD_stat,3)} for panel
d, and \Sexpr{round(panE_stat,3)} for panel e. These values were both statistically different from 
$0$ according to one-tailed tests based on an appropriate null hypothesis, $p=$\Sexpr{round(panD_pval,4)}
and $p=$\Sexpr{round(panE_pval,4)}, respectively. See below for details of the test.

Given scatterplot data, $(x_i,y_i)$ for $i=1,\ldots,N$, \cite{Ghosh_2020_AER} defined the 
\emph{normalized rank} $u_i$ of $x_i$ as the rank of $x_i$ in the set 
$\{x_1,\ldots,x_N\}$ divided by $N+1$, where the smallest value is taken to have rank $1$. 
The normalized ranks $v_i$ were defined for $y_i$ in a likewise manner. 
Then, the \emph{partial
Spearman correlation} for the left halves of the distributions, $\cor_{\text{left}}$, was defined as 
\begin{equation}
\frac{\sum(u_i-\mean(u))(v_i-\mean{v})}{(N-1)\sqrt{\var(u)\var(v)}},
\end{equation}
where the sum is over all indices $i$ for which $u_i+v_i<1$ (though the means and variances
in the above formula are computed over all the data). The indices satisfying $u_i+v_i<1$
are those corresponding to points falling under the dashed diagonal lines in Fig. \ref{MT-fig:pedagog1}d-e.
%***DAN: Note I mentioned a feature which Jasmin has not yet added, so make sure she 
%adds it to those panels.
The partial Spearman correlation for the right halves of the distributions,
$\cor_{\text{right}}$, was defined using the same formula but where the sum is 
now over all indices $i$ for which $u_i+v_i>1$. These are the points falling above the
dashed diagonal lines in Fig. \ref{MT-fig:pedagog1}d-e. The left and right partial Spearman
correlations typically add up to the Spearman correlation. The index 
$\cor_{\text{left}}-\cor_{\text{right}}$ is the index used by \cite{Ghosh_2020_AER}
mentioned in the previous paragraph.

To calculate whether the index of \cite{Ghosh_2020_AER} was significantly positive for 
Fig. \ref{MT-fig:pedagog1}d,
we compared the index value computed for the data of that panel to values computed for an appropriate 
symmetric distribution. We used the function `iRho' in the R package `copula'
to determine that the Spearman correlation of a bivariate normal distribution with
standard normal marginals and covariance parameter \Sexpr{round(panD_parm,4)}
is the same as the Spearman correlation of the data in Fig. \ref{MT-fig:pedagog1}d.
We simulated \Sexpr{nsims} data sets from such a distribution, each of size the same as 
that of the data set pictured in Fig. \ref{MT-fig:pedagog1}d. For each simulated data set,
we computed the index $\cor_{\text{left}}-\cor_{\text{right}}$ of \cite{Ghosh_2020_AER}.
The $p$-value quoted above is the fraction of values so obtained that were 
greater than the value $\cor_{\text{left}}-\cor_{\text{right}}$=\Sexpr{round(panD_stat,3)}
obtained for the real data. A similar approach was used for Fig. \ref{MT-fig:pedagog1}e.
